Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-20 main.tex
rule permutation

%All statements in this section are proved by rule permutation arguments.
\begin{thm}\label{thm:LK1-decompose}
  Let $\Gamma$ be a sequent. If\/ $\sqn\Gamma$ is provable in $\FOLK$ (as depicted on the left below) then there is a sequent
  $\Gamma'$ such that there is a derivation as shown on the right below:\vadjust{\vskip-2ex}
  \begin{equation*}
    \vlderivation{
        \vltrl{}{\FOLK}{\Deri}{\sqns{\Gamma}}{
          \vlhy{\quad}}{
          \vlhy{}}{
          \vlhy{}}}
    \qquad
    \Longrightarrow
    \qquad
    \vlderivation{
      \vlde{\set{\wrD,\cD,\fequ}}{\Deri_2}{\sqns{\form{\Gamma}}}{
        \vltrl{}{\FOMLL}{\Deri_1}{\sqns{\Gamma'}}{
          \vlhy{\quad}}{
          \vlhy{}}{
          \vlhy{}}}}
  \end{equation*}
\end{thm}
\juihsuan{maybe there is still some problem... (consider the proof

$\vlderivation{
   \vliin{\vlan}{}{\sqn{\ttt, a \vlan b, \cneg b}}{
   \vlin{\ax}{}{\sqn{b, \cneg b}}{
     \vlhy{}}}{
   \vlin{\wk}{}{\sqn{\ttt, a}}{
     \vlin{\ttt}{}{\sqn\ttt}{
       \vlhy{}}}}}
$, it doesn't seem possible to get rid of the $\wk$ rule using deep weakening)}
\begin{proof}
  Note that the instances of $\wrD,\cD$ in $\Deri_2$ are deep, but inside sequent contexts.
  Suppose that $\Gamma$ is provable in $\FOLK$, then so is $\form{\Gamma}$.
Consider a proof
$ \vlderivation{
        \vltrl{}{\FOLK}{\Deri}{\sqns{\form{\Gamma}}}{
          \vlhy{\quad}}{
          \vlhy{}}{
          \vlhy{}}} $
Note that the instances of $\ctr$ and $\wk$ in $\Deri$ can be transformed into
derivations using $\cD$, $\wrD$ and the rules from $\FOMLL$.

\begin{equation*}
\vlderivation{
  \vlin{\ctr}{}{\sqns{\Gamma, A}}{
    \vlhy{\sqns{\Gamma, A, A}}}}
\leadsto
\vlderivation{
  \vlin{\cD}{}{\sqns{\Gamma, A}}{
    \vlin{\vlor}{}{\sqns{\Gamma, A \vlor A}}{
      \vlhy{\sqns{\Gamma, A, A}}}}}
\end{equation*}

\todo{}
\end{proof}

>>>>

%All statements in this section are proved by rule permutation arguments.
\begin{thm}\label{thm:LK1-decompose}
  Let $\Gamma$ be a sequent. If\/ $\sqn\Gamma$ is provable in $\FOLK$ (as depicted on the left below) then there is a sequent
  $\Gamma'$ such that there is a derivation as shown on the right below:\vadjust{\vskip-2ex}
  \begin{equation*}
    \vlderivation{
        \vltrl{}{\FOLK}{\Deri}{\sqns{\Gamma}}{
          \vlhy{\quad}}{
          \vlhy{}}{
          \vlhy{}}}
    \qquad
    \Longrightarrow
    \qquad
    \vlderivation{
      \vlde{\set{\wrD,\cD,\fequ}}{\Deri_2}{\sqns{\form{\Gamma}}}{
        \vltrl{}{\FOMLL}{\Deri_1}{\sqns{\Gamma'}}{
          \vlhy{\quad}}{
          \vlhy{}}{
          \vlhy{}}}}
  \end{equation*}
\end{thm}

\begin{proof}

  Note that the instances of $\wrD,\cD$ in $\Deri_2$ are deep, but inside sequent contexts.

  Suppose that $\Gamma$ is provable in $\FOLK$, then so is $\form{\Gamma}$.
Consider a proof
$ \vlderivation{
        \vltrl{}{\FOLK}{\Deri}{\sqns{\form{\Gamma}}}{
          \vlhy{\quad}}{
          \vlhy{}}{
          \vlhy{}}} $

First, if an instance $\vlinf{\weakr}{}{\sqn{\Gamma,A}}{\sqn{\Gamma}}$
 of $\weakr$ is followed by a rule in which $A$ is not in the principal
formula, it can be permuted downwards.
Otherwise, the proof can be transformed using the following rewriting rules.

\begin{equation*}
\vlderivation{
  \vliin{\vlan}{}{\sqns{\Gamma, A \vlan B, \Delta}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A}}{
      \vlhy{\sqns{\Gamma}}}}{
    \vlhy{\sqns{B, \Delta}}}}
\leadsto
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, A \vlan B, \Delta}}{
    \vlin{\weakr}{}{\vlvdots}{
      \vlhy{\sqns{\Gamma}}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\vlor}{}{\sqns{\Gamma, A \vlor B}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A, B}}{
      \vlhy{\sqns{\Gamma, A}}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{\sqns{\Gamma, A \vlor B}}{
    \vlhy{\sqns{\Gamma, A}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\exists}{}{\sqns{\Gamma, \exists x.A}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A\ssubst{x}{t}}}{
      \vlhy{\sqns{\Gamma}}}}}
\leadsto
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, A\ssubst{x}{t}}}{
    \vlhy{\sqns{\Gamma}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\forall}{}{\sqns{\Gamma, \forall x.A}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A}}{
      \vlhy{\sqns{\Gamma}}}}}
\leadsto
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, \forall x.A}}{
    \vlhy{\sqns{\Gamma}}}}
\end{equation*}
\begin{equation*}
\vlderivation{
  \vlin{\conr}{}{\sqns{\Gamma, A}}{
    \vlin{\weakr}{}{\sqns{\Gamma, A, A}}{
      \vlhy{\sqns{\Gamma, A}}}}}
\leadsto
\vlderivation{
  \vlhy{\sqns{\Gamma, A}}}
\end{equation*}
Note that in the case of $\vlor$, we use the deep rule $\wrD$ which can be
permuted down over all the rules. By using these rewriting rules, we can eventually get
a derivation with all the instances of $\weakr$ and $\wrD$ at the bottom. Now observe
that the instances of $\conr$ in $\Deri$ can be transformed using the following
rule:

\begin{equation*}
\vlderivation{
  \vlin{\conr}{}{\sqns{\Gamma, A}}{
    \vlhy{\sqns{\Gamma, A, A}}}}
\leadsto
\vlderivation{
  \vlin{\cD}{}{\sqns{\Gamma, A}}{
    \vlin{\vlor}{}{\sqns{\Gamma, A \vlor A}}{
      \vlhy{\sqns{\Gamma, A, A}}}}}
\end{equation*}

Knowing that $\cD$ can be permuted down over all the rules of $\FOMLL$, we
eventually obtain a derivation:
\begin{equation*}
\vlderivation{
  \vlde{\set{\weakr, \wrD,\cD,\fequ}}{\Deri'_2}{\sqns{\Gamma}}{
    \vltrl{}{\FOMLL}{\Deri'_1}{\sqns{\Gamma_0}}{
      \vlhy{\quad}}{
      \vlhy{}}{
      \vlhy{}}}}
\end{equation*}
Note that $\fequ$ is required here since the permutaion of formulas is implicit
in $\FOMLL$.

By transforming each sequent of $\Deri'_2$ into its corresponding formula, and by
considering the following rewriting rule:
\begin{equation*}
\vlderivation{
  \vlin{\weakr}{}{\sqns{\Gamma, A}}{
    \vlhy{\sqns{\Gamma}}}}
\leadsto
\vlderivation{
  \vlin{\wrD}{}{\sqns{\form{\Gamma} \vlor A}}{
    \vlhy{\sqns{\form{\Gamma}}}}}
\end{equation*},
we obtain a derivation
\begin{equation*}
\vlderivation{
  \vlde{\set{\wrD,\cD,\fequ}}{\Deri_2}{\sqns{\form{\Gamma}}}{
    \vltrl{}{\FOMLL}{\Deri_1}{\sqns{\Gamma'}}{
      \vlhy{\quad}}{
      \vlhy{}}{
      \vlhy{}}}}
\end{equation*}
where $\Gamma' = \form{\Gamma_0}$ and $\Deri_1$ can be obtained from $\Deri'_1$
by applying the $\vlor$ rule.
\tocheck{}
\juihsuan{This might be a bit long...}
\end{proof}
